---
title: "Exploratory analysis"
author: "João Gabriel Solar"
date: "`r Sys.Date()`"
output: github_document
---

## Load libraries and functions
```{r, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
library(tidyverse)
library(here)
source(here("C_scrub","a_R_tidy_verse_data_engineering.R"))
```

## Collect data
To increment available data, I'm including 
[horse-survival-dataset](https://www.kaggle.com/datasets/yasserh/horse-survival-dataset)
in the training set.

```{r, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
data <- read_csv(here("A_data", "train.csv"))
data_orig <- read_csv(here("A_data", "horse.csv"))
data <- data %>% select(!c('id'))
data <- data %>% rbind(data_orig)
data <- data %>% drop_na('outcome')
rm(data_orig)
```

## Attributes analysis
To evaluate data tidiness, I'm using [horse-survival-dataset](https://www.kaggle.com/datasets/yasserh/horse-survival-dataset)
features specifications.


```{r}
data <- data %>%
  mutate_if(is.character, as.factor)

summary(data)
```
There are several categorical attributes with wrong coding. I did not find any 
anomalies in the quantitative attributes. `cp_data` doesn't have any relevant
information, according to reference [horse-survival-dataset](https://www.kaggle.com/datasets/yasserh/horse-survival-dataset)
and will be removed.

## Lesions codification

The lesion code doesn't respect normal form and must be separated. I'm using the
following criteria to address this separation:
1. Counting the number of lesions and recording it in a new feature `numLesions`.  
2. During the process, I observed that a few horses had more than one lesion.
Specifically, eight horses had two, and only two had three lesions. To simplify 
the analysis, I chose to consider only one lesion per horse, selecting the lesion
with the highest `lesionSite` code. (This strategy couldn't be applied in all
platforms, given the limitation and restrictions founded)  
3. This approach simplified feature engineering process, especially considering
the short data available.  
4. Finally, I decoded the lesions into five different variables: `numLesions`,
`lesionSite`, `lesionType`, `lesionSubType` and `lesionCode`, each of them 
representing one dimension of lesion encoded.

## Data cleaning

For better comprehension of data characteristics, I'll perform with data 
cleaning before proceeding with exploratory analysis.

```{r, results='hide', message=FALSE, warning=FALSE}
data <- data %>%
  mutate_if(is.factor, as.character)  # revertion of previous operation
results = scrub(data, train=TRUE)
data <- results[[1]]
metadata <- results [[2]]
rm(results)
```


## Exploratory analysis

Categorical attributes histogram analysis:

```{r, echo=TRUE, results='hold', message=FALSE, warning=FALSE, fig.keep="all"}
outcome = data['outcome']
outcome <- outcome %>% mutate(outcome =
                              gsub("2", "died", outcome))
outcome <- outcome %>% mutate(outcome =
                          gsub("1", "euthanized", outcome))
outcome <- outcome %>% mutate(outcome =
                          gsub("0", "lived", outcome))

cat_cols <- metadata$cat_cols
data_temp <- data[cat_cols]

data_temp['outcome'] <- outcome

histograms <- lapply(names(data_temp), function(var_name) {
  p <- ggplot(data_temp, aes(x = .data[[var_name]])) +
    geom_bar() +
    labs(title = var_name)
})

for (hist in histograms) {
  print(hist)
}

rm(data_temp, outcome)
```

There are imbalanced data in categorical attributes, but I'll keep them as
presented originally. The outcome variable, when possible, will pass 
through a down sample process to balance it.

```{r, fig.keep="all"}
num_cols <- metadata$num_cols

histograms <- lapply(names(data[c(num_cols)]), function(var_name) {
  p <- ggplot(data, aes(x = .data[[var_name]])) +
    geom_histogram() +
    labs(title = var_name)
})

for (hist in histograms) {
  print(hist)
}
```



By the distribution of histogram, the variables `abdomo_protein`, `pulse`, 
`respiratory_rate`, `total_protein` are left skewed. When possible, they will 
be treated in the pipeline. In tidy models, as an ORQ Normalization 
(step_orderNorm) is available, all numerical attributes will pass through it.

```{r}
left_skewed = c('abdomo_protein', 'pulse', 'respiratory_rate',
               'total_protein')
```


```{r, echo=TRUE, results='hold', message=FALSE, warning=FALSE, fig.keep="all"}

outcome <- data$outcome %>% as.numeric()
data_num <- cbind(data[c(num_cols)], outcome)

# Correlation verification
matrix_cor <- cor(data_num[, 1:ncol(data_num)], use = "complete.obs")

# Check correlation through heatmap
matrix_cor_melt <- reshape2::melt(matrix_cor)
ggplot(matrix_cor_melt, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Correlation Matrix")
```

Any variable were found with great correlation with the outcome. I decided 
not to remove any variables, even if they have a low correlation with 
the outcome, because combined variables can be significant, even if they aren't
individually significant. Variables with high correlation with each other
will be removed in the pipeline.



